I have developed a real-time continuous gesture recognition system that
addresses problems that have previously been neglected: handling both gestures that are
characterized by distinct paths and gestures characterized by distinct hand
poses; and determining how and when the system should respond to gestures.
The novel approaches in this thesis include: a probabilistic recognition
framework based on a flattened hierarchical hidden Markov model (HHMM) that
unifies the recognition of path and pose gestures;
a method of using information from the
hidden states in the HMM to identify different
gesture phases (the pre-stroke, the nucleus and the post-stroke
phases), allowing the system to respond appropriately to both gestures that
require a discrete response and those needing a continuous response.

I collected a new dataset (YANG dataset) that includes two forms of
gestures (4 path and discrete flow gestures and 3 pose and continuous flow
gestures from 10 users), a combination currently lacking in the community, to
evaluate system performance. This is a contribution to the gesture recognition
community because it presents a challenge of recognizing different
types of gestures mixed together. I also developed a novel hybrid evaluation
metric that is more relevant to real-time interaction with different gesture flows.

The system is extensible: new gestures can be added by recording 3-6 repetitions
of the gesture using the Kinect; the system will train an HMM model for the gesture
and add it to the existing combined HMM. This process takes only a few minutes.
The evaluation shows that even using only a small number of
training examples (e.g. 6), the system can achieve an average $F_1$ score of 0.805 for
two forms of gestures.
