\begin{savequote}
Language is a part of social behavior. What is the mechanism whereby the social
process goes on? It is the mechanism of gesture\ldots
\qauthor{George Herbert Mead, \textit{Mind, self, and society}}
\end{savequote}
\chapter{Conclusion}
I developed a real-time continuous gesture recognition system for natural
human computer interaction. The
unified probabilistic framework for handling two forms of gestures is a novel
approach, and the evaluation shows promising results: an
average online recognition $F_1$ score of 0.805 using the hybrid performance
metric, on a challenging dataset with unsegmented gestures of different forms.
The system also achieves real-time performance at 30FPS. Using the framework, I
developed a gesture controlled presentation application similar to the one described at the beginning of this paper. All the code is open-source
and is available
online\footnote{\url{http://groups.csail.mit.edu/mug/projects/gesture_kinect/index.html\#code}}.

Another novel approach is using embedded training and hidden state information
to detect gesture phases, allowing the system to respond more promptly. On
average, for discrete flow gestures, the system responds 0.6s before the hand comes to
rest. 

I collected a new dataset that includes two forms of gestures, a
combination currently lacking in the community, and plan to make it
public. I also proposed a hybrid performance metric
that is more relevant to real-time interaction and different types of gestures.

\section{Limitations}
Currently, the system only handles single-hand gestures, it will not be hard to
add two-hand gestures. The only difference will be the feature vector which
should be a concatenation of features from both hands. The recognition module
would remain the same.

The number of hidden states per gesture needs to be specified by the user or the
application developer in the gesture definition file. In the future work, we
will make the system automatically learn the number of hidden states through
cross-validation and by grouping similar frames~\cite{song13}.

The performance for pose gestures is lower than that
for path gestures. Two factors contribute to this problem: the limitation in
both the pixel and the depth resolutions of the Kinect sensor, and motion blur.
It would be interesting to test with the new version of the Kinect sensor which
uses a time-of-flight depth sensor and is reported to have a higher resolution. Other feature descriptors
(e.g., histograms of optic flows) and encoding methods (e.g, sparse
coding~\cite{lee07}) can be explored as well.

The performance of a user independent model is relatively low. This is expected
because there is large variation among users. Even though, users can
quickly define and train their own models, an relatively accurate base
model is always valuable. Hence, more work can be done to improve the base
model, e.g., using mixture of HMMs~\cite{keskin12}.

\section{Future Work}
In addition to work that address the limitations. There is future research to
consider to push the frontier of natural human computer interaction further.

\textbf{Sensors} In this thesis, I considered the Kinect and the IMU as
input sensors. The gesture controlled presentation application currently
only uses the Kinect data. We can explore even more combinations of sensors
by including the Leap Motion Controller and smartwatches. Each sensor has
strengths of their own, so by combining them we can get the best system
overall. The Leap Motion Controller is mostly used as an environmental sensor,
but we can also explore its use case a wearable sensor. Its high accuracy in
finger tracking may help improve recognition performance for pose gestures.

\textbf{User adaptation} This is a hot topic in both machine learning and
HCI, and is applicable to many domains such as speech recognition and
touch screen keyboard input~\cite{yin13-making}. Currently, the system
uses a binary decision to do user adaptation: if a user-trained model is
present, the system will use that model; otherwise it uses the base
model. It would be more convenient if the system learns the user
dependent model implicitly when the user is using the system, and uses
soft decision to combine the base model and user dependent model, e.g., weighted
combination.

\textbf{Interactive training} Due to large variation in users performing
gestures and difference in user preference, the gesture recognition system would
benefit from having an easy to use interface for fast training. Many machine
learning related research assumes the availability of training data. Creating
interfaces that are easy for users to provide training examples, either in a
separate session, or when using the system and provide feedback to the system
for wrong recognition would be an interest interdisciplinary topic in machine
learning and HCI.
